---
title: "transform_kuf_nas"
author: "Nico Valencia"
date: "2025-10-05"
output: html_document
---

The goal of this script is to transform NAs in Khat Use Frequency to 0 if their assist_khat value = 0. These new recoded "0" values correspond to people who have never used in their life. This differentiates from the individuals who "Have not used in the past 3 months" (assist_khat_amt = 1).
  - In this process, one extra sample at UCT was dropped due to not passing the aforementioned argument (i.e., still remained an NA, so we dropped it)

I also added a section on getting stacked barplots for distribution of khat use and khat use frequency across each site.

```{r}
# Load library
library(ggplot2)
library(dplyr)
library(scales)
```

```{r}
# read in phenotype files
w_site_Uganda_pheno_noNA <- read.csv("~/Desktop/work/khat_gwas/script_referenced_files/gp08_covar_adj/w_site_Uganda_pheno_noNA.txt", sep="")

w_site_AAU_pheno_noNA <- read.csv("~/Desktop/work/khat_gwas/script_referenced_files/gp08_covar_adj/w_site_AAU_pheno_noNA.txt", sep="")

w_site_UCT_pheno_noNA <- read.csv("~/Desktop/work/khat_gwas/script_referenced_files/gp08_covar_adj/w_site_UCT_pheno.txt", sep="")

w_site_KEMRI_pheno_noNA <- read.csv("~/Desktop/work/khat_gwas/script_referenced_files/gp08_covar_adj/w_site_KEMRI_pheno.txt", sep="")

w_site_Moi_pheno_noNA <- read.csv("~/Desktop/work/khat_gwas/script_referenced_files/gp08_covar_adj/w_site_Moi_pheno.txt", sep="")

```


```{r}
# Uganda
w_site_Uganda_pheno_noNA$assist_khat_amt[
  w_site_Uganda_pheno_noNA$assist_khat == 0 & is.na(w_site_Uganda_pheno_noNA$assist_khat_amt)
] <- 0

# AAU
w_site_AAU_pheno_noNA$assist_khat_amt[
  w_site_AAU_pheno_noNA$assist_khat == 0 & is.na(w_site_AAU_pheno_noNA$assist_khat_amt)
] <- 0

# UCT
w_site_UCT_pheno_noNA$assist_khat_amt[
  w_site_UCT_pheno_noNA$assist_khat == 0 & is.na(w_site_UCT_pheno_noNA$assist_khat_amt)
] <- 0

# KEMRI
w_site_KEMRI_pheno_noNA$assist_khat_amt[
  w_site_KEMRI_pheno_noNA$assist_khat == 0 & is.na(w_site_KEMRI_pheno_noNA$assist_khat_amt)
] <- 0

# Moi
w_site_Moi_pheno_noNA$assist_khat_amt[
  w_site_Moi_pheno_noNA$assist_khat == 0 & is.na(w_site_Moi_pheno_noNA$assist_khat_amt)
] <- 0

```

```{r}
# show distribution of khat use frequency
show_dist <- function(df, site_name) {
  cat("\n---", site_name, "---\n")
  print(table(df$assist_khat_amt, useNA = "ifany"))
}

# print value distributions for each site
show_dist(w_site_Uganda_pheno_noNA, "Uganda")
show_dist(w_site_AAU_pheno_noNA, "AAU")
show_dist(w_site_UCT_pheno_noNA, "UCT")
show_dist(w_site_KEMRI_pheno_noNA, "KEMRI")
show_dist(w_site_Moi_pheno_noNA, "Moi")

```


```{r}
# remove any NAs samples w/ NAs

w_site_Uganda_pheno_noNA <- subset(w_site_Uganda_pheno_noNA, !is.na(assist_khat_amt))
w_site_AAU_pheno_noNA    <- subset(w_site_AAU_pheno_noNA,    !is.na(assist_khat_amt))
w_site_UCT_pheno_noNA    <- subset(w_site_UCT_pheno_noNA,    !is.na(assist_khat_amt))
w_site_KEMRI_pheno_noNA  <- subset(w_site_KEMRI_pheno_noNA,  !is.na(assist_khat_amt))
w_site_Moi_pheno_noNA    <- subset(w_site_Moi_pheno_noNA,    !is.na(assist_khat_amt))

# show distribution of khat use frequency
show_dist <- function(df, site_name) {
  cat("\n---", site_name, "---\n")
  print(table(df$assist_khat_amt, useNA = "ifany"))
}

# print updated distributions
show_dist(w_site_Uganda_pheno_noNA, "Uganda")
show_dist(w_site_AAU_pheno_noNA, "AAU")
show_dist(w_site_UCT_pheno_noNA, "UCT")
show_dist(w_site_KEMRI_pheno_noNA, "KEMRI")
show_dist(w_site_Moi_pheno_noNA, "Moi")


```

```{r}
### MAKE STACKED BARPLOT FOR KHAT USE FREQUENCY

# Combine all sites into one dataframe
w_site_Uganda_pheno_noNA$site <- "Uganda"
w_site_AAU_pheno_noNA$site    <- "Ethiopia"
w_site_UCT_pheno_noNA$site    <- "South Africa"
w_site_KEMRI_pheno_noNA$site  <- "KEMRI"
w_site_Moi_pheno_noNA$site    <- "Moi"

combined_df <- bind_rows(
  w_site_Uganda_pheno_noNA,
  w_site_AAU_pheno_noNA,
  w_site_UCT_pheno_noNA,
  w_site_KEMRI_pheno_noNA,
  w_site_Moi_pheno_noNA
)

# Convert to numeric if needed
combined_df$assist_khat_amt <- as.numeric(as.character(combined_df$assist_khat_amt))

# Map numeric codes to descriptive labels
label_map <- c(
  "0" = "Never Used",
  "1" = "Have not Used",
  "2" = "Once or Twice",
  "3" = "Monthly",
  "4" = "Weekly",
  "5" = "Daily or Almost Daily"
)

combined_df$assist_khat_amt_label <- factor(
  label_map[as.character(combined_df$assist_khat_amt)],
  levels = c(
    "Daily or Almost Daily",
    "Weekly",
    "Monthly",
    "Once or Twice",
    "Have not Used",
    "Never Used"
  ),
  ordered = TRUE
)

# Summarize counts
count_df <- combined_df %>%
  group_by(site, assist_khat_amt_label) %>%
  summarise(count = n(), .groups = "drop")

# Bright blue palette (darker = higher frequency)
levs <- levels(count_df$assist_khat_amt_label)
pal  <- setNames(colorRampPalette(c("#1565C0", "#BBDEFB"))(length(levs)), levs)

# Plot (will stack 5→0 visually if you keep position_stack(reverse = TRUE))
ggplot(count_df, aes(x = site, y = count, fill = assist_khat_amt_label)) +
  geom_bar(stat = "identity", color = "white", position = position_stack(reverse = TRUE)) +
  scale_fill_manual(
    values = pal,
    name = "Usage Past 3 Months",
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    title = "Distribution of Khat Use Frequency in 3 Months",
    x = "Study Site",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

```

```{r}
## --- Percent labels for "Have used" vs "Never Used" ---
site_totals <- count_df %>%
  group_by(site) %>%
  summarise(total = sum(count), .groups = "drop")

never_counts <- count_df %>%
  filter(assist_khat_amt_label == "Never Used") %>%
  select(site, never_count = count)

labels_df <- site_totals %>%
  left_join(never_counts, by = "site") %>%
  mutate(
    never_count = dplyr::coalesce(never_count, 0L),
    have_count  = total - never_count,
    label_never = percent(never_count / total, accuracy = 0.1),
    label_have  = percent(have_count  / total, accuracy = 0.1),
    y_never     = total,       # top of full bar (Never Used segment)
    y_have      = have_count   # boundary between Have used and Never Used
  )
```

```{r}

# Combine all site data
w_site_Uganda_pheno_noNA$site <- "Uganda"
w_site_AAU_pheno_noNA$site    <- "Ethiopia"
w_site_UCT_pheno_noNA$site    <- "South Africa"
w_site_KEMRI_pheno_noNA$site  <- "KEMRI"
w_site_Moi_pheno_noNA$site    <- "Moi"

combined_df <- bind_rows(
  w_site_Uganda_pheno_noNA,
  w_site_AAU_pheno_noNA,
  w_site_UCT_pheno_noNA,
  w_site_KEMRI_pheno_noNA,
  w_site_Moi_pheno_noNA
)

# Recode assist_khat (0/1) to descriptive labels
combined_df <- combined_df %>%
  mutate(
    assist_khat_label = factor(
      ifelse(assist_khat == 1, "Have used", "Never used"),
      levels = c("Have used", "Never used")
    )
  )

# Summarize counts and percentages
count_df <- combined_df %>%
  group_by(site, assist_khat_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(site) %>%
  mutate(
    total = sum(count),
    percent = 100 * count / total
  ) %>%
  ungroup() 

# Compute y positions for labels
label_df <- count_df %>%
  group_by(site) %>%
  arrange(site, assist_khat_label) %>%
  mutate(
    cum_count = cumsum(count),
    y_label = case_when(
      assist_khat_label == "Have used" ~ cum_count + 250,   
      assist_khat_label == "Never used" ~ cum_count + 250   
    )
  )


# Define colors
pal <- c("Have used" = "#1565C0", "Never used" = "#BBDEFB")

# --- Plot ---
ggplot(count_df, aes(x = site, y = count, fill = assist_khat_label)) +
  geom_bar(stat = "identity", color = "white", position = position_stack(reverse = TRUE)) +
  # Add percentage labels (now with explicit x mapping)
  geom_text(
    data = label_df,
    aes(x = site, y = y_label, label = paste0(round(percent, 1), "%")),
    inherit.aes = FALSE,
    size = 4,
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = pal,
    name = "Khat Use",
    guide = guide_legend(reverse = TRUE)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Distribution of Khat Use by Site",
    x = "Study Site",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )


```


```{r}

# export files
# Define output directory
out_dir <- "~/Desktop/work/khat_gwas/script_referenced_files/gp08_covar_adj/"

# Write each dataframe to file
write.table(
  w_site_Uganda_pheno_noNA,
  file = paste0(out_dir, "kuf_NAs0_w_site_Uganda_pheno.txt"),
  sep = "\t", row.names = FALSE, quote = FALSE
)

write.table(
  w_site_AAU_pheno_noNA,
  file = paste0(out_dir, "kuf_NAs0_w_site_AAU_pheno.txt"),
  sep = "\t", row.names = FALSE, quote = FALSE
)

write.table(
  w_site_UCT_pheno_noNA,
  file = paste0(out_dir, "kuf_NAs0_w_site_UCT_pheno.txt"),
  sep = "\t", row.names = FALSE, quote = FALSE
)

write.table(
  w_site_KEMRI_pheno_noNA,
  file = paste0(out_dir, "kuf_NAs0_w_site_KEMRI_pheno.txt"),
  sep = "\t", row.names = FALSE, quote = FALSE
)

write.table(
  w_site_Moi_pheno_noNA,
  file = paste0(out_dir, "kuf_NAs0_w_site_Moi_pheno.txt"),
  sep = "\t", row.names = FALSE, quote = FALSE
)

```


